Joint maximum-likelihood of phylogeny and ancestral states is not consistent

### 2018-08-06 joint_inf.tex
some restructuring mostly

\title{Joint maximum-likelihood of phylogenies and ancestral states is not consistent}

@@

Our main theorem states that given data generated on $\tau_1$ there exist true nonzero branch lengths such that the estimator $\hat{t}$ maximizing the right-hand side of \eqref{eq:exp-log-lik} has an internal branch of length zero.
%EM A funny thing: this theorem as stated is really under the hypothesis that the ancestral state is emptyset. We can combine this with the lemmas to remove this assumption, but it seems like we should do this for the reader. Time for a corollary?
\begin{restatable}{theorem}{topoInconsist}
\label{thm:main_theorem}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
%EM Can we say "open set" or "set of positive measure"?
For $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$ there exists a set of $0 < x^*, y^* < 1$ such that the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_{t} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi_0)
\]
has the property $\hat{w}\equiv 1$.
\end{restatable}

This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one).
As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent.
We elaborate on this point in the Appendix.
The proof is through analyzing upper and lower bounds of the potential conditional likelihood forms in Table~\ref{tab:farris_likelihoods} and the gradients of the objective function $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ with respect to $t$.

@@

Previous work in phylogenetics has developed consistency counterexamples using similar four-taxon topologies to the one used here \citep{Felsenstein1978-rr}.
In this previous work, when simulating data under the Felsenstein topology $\tau_2$, as the number of observations increases, the InvFels topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
We have shown cases in which, when generating from the InvFels topology, we converge to a multifurcating topology, with one or more branch lengths estimated to be zero.
Moreover, the inconsistency demonstrated by \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
%EM I don't follow here. This doesn't seem like the long branch attraction region of parameter space-- isn't that on the Felsenstein tree, not the InvFels tree?
%das: yeah, it was kind of hand-wavy.
This is not the case here; while analytically the inconsistency occurs when the top two branches are long and the bottom three are short, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
%EM Is this next sentence necessary given the previous one?
This is especially apparent in the ``long $x^*$, short $y^*$'' case discussed previously.
Additionally, we generate data on the InvFels tree $\tau_1$ while \citet{Felsenstein1978-rr} generates data on the Felsenstein tree $\tau_2$.
Difficulties in phylogenetic estimation when generating data on the InvFels tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.

>>>>

\title{Joint maximum-likelihood of phylogeny and ancestral states is not consistent}

@@

Our main theorem states that given data generated on $\tau_1$ there exist true nonzero branch lengths such that the estimator $\hat{t}$ maximizing the right-hand side of \eqref{eq:exp-log-lik} has an internal branch of length zero.
%EM A funny thing: this theorem as stated is really under the hypothesis that the ancestral state is emptyset. We can combine this with the lemmas to remove this assumption, but it seems like we should do this for the reader. Time for a corollary?
%das: ok, I reworded it a bit; let me know how you think it sounds.
\begin{restatable}{theorem}{topoInconsist}
\label{thm:main_theorem}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
%EM Can we say "open set" or "set of positive measure"?
There exists an open set of $0 < x^*, y^* < 1$ such that for some $\boldsymbol\xi$ the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_{t} \ \ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)
\]
has the property $\hat{w}\equiv 1$.
\end{restatable}

This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:incons-analytic}).
As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent.
We elaborate on this point in the Appendix.
The proof is through analyzing upper and lower bounds of the potential conditional likelihood forms in Table~\ref{tab:farris_likelihoods} and the gradients of the objective function $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ with respect to $t$.

@@

Previous work in phylogenetics has developed consistency counterexamples using similar four-taxon topologies to the one used here \citep{Felsenstein1978-rr}.
In this previous work, when simulating data under the Felsenstein topology $\tau_2$, as the number of observations increases, the InvFels topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
We have shown cases in which, when generating from the InvFels topology, we converge to a multifurcating topology, with one or more branch lengths estimated to be zero.
Moreover, the inconsistency demonstrated by \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
%EM I don't follow here. This doesn't seem like the long branch attraction region of parameter space-- isn't that on the Felsenstein tree, not the InvFels tree?
%das: yeah, it was kind of hand-wavy.
This is not the case here; while analytically the inconsistency occurs when the top two branches are long and the bottom three are short, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
%EM Is this next sentence necessary given the previous one?
Additionally, we generate data on the InvFels tree $\tau_1$ while \citet{Felsenstein1978-rr} generates data on the Felsenstein tree $\tau_2$.
Difficulties in phylogenetic estimation when generating data on the InvFels tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.
